Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-08 _tex/index.tex
Built site for gh-pages

In \emph{comparative judgment} (CJ) studies, judges assess the presence
of a trait or competence in different stimuli by performing pairwise
comparisons \citep{Thurstone_1927, Pollitt_2004, Pollitt_2012a}. The
comparison produces a dichotomous outcome, indicating which stimulus is
perceived to possess a higher trait level. After multiple rounds of
pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} to process the resulting outcomes
and estimate scores that reflect the underlying trait of interest. This
method has been successfully employed in assessing the quality of
written texts, where quality describes the underlying trait of interest
and the texts serve as the stimuli
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

Numerous studies have documented the effectiveness of CJ in assessing
various traits and competencies over the past decade. These studies have
emphasized three aspects of the method's effectiveness: its reliability,
validity, and practical applicability. Research on reliability indicates
that CJ requires a relatively small number of pairwise comparisons
\citep{Verhavert_et_al_2019, Crompvoets_et_al_2022} to produce trait
scores that are as precise and consistent as those generated by other
assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
Furthermore, evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
Additionally, research on validity suggests that scores generated by CJ
can accurately represent the traits under measurement
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
Finally, research on practical applicability highlights the method's
versatility across both educational and non-educational contexts
\citep{Jones_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.

Nevertheless, despite the growing number of CJ studies, the literature's
unsystematic and fragmented research approaches have left several
critical issues unaddressed. This research primarily focuses on three:
the apparent disconnect between CJ's measurement and structural model,
the over-reliance on the assumptions of Thurstone's Case 5
\citeyearpar{Thurstone_1927} in CJ's measurement model, and the unclear
role of comparison algorithms on the method's reliability and validity.
The following sections will discuss each of these issues in detail,
followed by the introduction of a theoretical model and its statistical
translation, which aims to address all three concerns simultaneously.

@@

In a typical CJ study, the BTL model serves as the measurement model for
CJ \citep{Andrich_1978, Bramley_2008}. A measurement model specifies how
manifest variables contribute to the estimation of latent variables
\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
the BTL model uses the dichotomous outcomes resulting from the pairwise
comparisons (the manifest variables) to estimate scores that reflect the
underlying quality level of the texts (the latent variable)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

@@

However, the statistical literature advises caution when using estimated
scores to conduct additional analyses or hypotheses tests. A key
consideration is that BTL scores are parameter estimates that inherently
carry uncertainty. Ignoring this uncertainty can introduce bias into the
analysis and reduce the precision of hypothesis tests. Notably, the
direction and magnitude of this bias are often unpredictable; results
may be attenuated, exaggerated, or remain unaffected, depending on the
uncertainty in the scores and the actual effects being tested
\citetext{\citealp[p.~25]{Kline_et_al_2023}; \citealp[p.~137]{Hoyle_et_al_2023}}.
Furthermore, reduced precision in hypothesis tests weakens their
statistical power, ultimately increasing the likelihood of committing
type-I or type-II errors \citep{McElreath_2020}.

To mitigate these risks, principles from Structural Equation Modeling
(SEM) \citep[p.~138]{Kline_et_al_2023, Hoyle_et_al_2023} and Item
Response Theory (IRT)
\citep{deAyala_2009, Fox_2010, vanderLinden_et_al_2017} recommend
conducting these analyses and tests within a structural model. A
structural model specifies how different manifest or latent variables
influence the latent variable of interest \citep{Everitt_et_al_2010}.
This approach allows analyses that can account for both the scores and
their uncertainties simultaneously, rather than treating them as
separate elements. Therefore, an integrated approach that combines CJ's
measurement and structural models can offer significant advantages.

>>>>

In \emph{comparative judgment} (CJ) studies, judges assess a specific
trait or competence across various stimuli by performing pairwise
comparisons \citep{Thurstone_1927, Pollitt_2004, Pollitt_2012a}. Each
comparison produces a dichotomous outcome, indicating which stimulus is
perceived to manifest a higher trait level. For example, when assessing
text quality, judges compare pairs of written texts (the stimuli) to
determine the relative quality each text exhibit (the trait)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

Numerous studies have documented the effectiveness of CJ in assessing
traits and competencies over the past decade. These studies have
emphasized three aspects of the method's effectiveness: its reliability,
validity, and practical applicability. Research on reliability indicates
that CJ requires a relatively small number of pairwise comparisons
\citep{Verhavert_et_al_2019, Crompvoets_et_al_2022} to produce trait
scores that are as precise and consistent as those generated by other
assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
Furthermore, evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}. On
the other hand, research on validity suggests that scores generated by
CJ can accurately represent the traits under measurement
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
Finally, research on practical applicability highlights the method's
versatility across both educational and non-educational contexts
\citep{Jones_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.

Nevertheless, despite the growing number of CJ studies, unsystematic and
fragmented research approaches in the literature have left several
critical issues unaddressed. This research primarily focuses on three:
the apparent disconnect between CJ's measurement and structural model,
the over-reliance on the assumptions of Thurstone's Case 5
\citeyearpar{Thurstone_1927} in CJ's measurement model, and the unclear
role of comparison algorithms on the method's reliability and validity.
The following sections will discuss each of these issues in detail,
followed by the introduction of a theoretical model and its statistical
translation, which aims to address all three concerns simultaneously.

@@

In a typical CJ study, the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} serves as the measurement model
for CJ \citep{Andrich_1978, Bramley_2008}. A measurement model specifies
how manifest variables contribute to the estimation of latent variables
\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
researchers use the BTL model to process the dichotomous outcomes
resulting from the pairwise comparisons (the manifest variables) to
estimate scores that reflect the underlying quality level of the texts
(the latent variable)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

@@

However, the statistical literature advises caution when using estimated
scores to conduct additional analyses or hypotheses tests. A key
consideration is that BTL scores are parameter estimates that inherently
carry uncertainty. Ignoring this uncertainty can introduce bias into the
analysis and reduce the precision of hypothesis tests. Notably, the
direction and magnitude of the bias are often unpredictable; results may
be attenuated, exaggerated, or remain unaffected, depending on the
amount of uncertainty present in the scores and the actual effects being
tested
\citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
Furthermore, reduced precision in hypothesis tests weakens their
statistical power, ultimately increasing the likelihood of committing
type-I or type-II errors \citep{McElreath_2020}.

To mitigate these risks, principles from Structural Equation Modeling
(SEM) \citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}}
recommend conducting these analyses and tests within a structural model.
A structural model specifies how different manifest or latent variables
influence the latent variable of interest \citep{Everitt_et_al_2010}.
This approach allows analyses that can account for both the scores and
their uncertainties simultaneously, rather than treating them as
separate elements. Therefore, an integrated approach that combines CJ's
measurement and structural models can offer significant advantages.
